---
title: "Reproducible Research - Assignment 1"
author: "Ali Belkairous"
date: "Tuesday, July 08, 2014"
output: html_document
---
**Data Management and Transformation**
```{r, setoptions, echo = TRUE}
setwd("c:/users/abelkair/RR")
act.df<-read.csv("c:/week4/activity.csv", header=TRUE)
act.df<-as.data.frame(act.df)
```
**Create a dataset with no missing values and with steps grouped by day**
```{r}
data <- aggregate(act.df$steps, list(Day=act.df$date), na.rm=TRUE, sum)
```
**Histogram of frequency of daily number of steps**

```{r histogram, fig.height=4}
names(data)[2]<-"Steps"
max_num <- max(data$Steps)
brk<-c(data$Day)
lim<-dim(data)[1]
data <- data[order(data$Steps),]

hist(data$Steps, col=heat.colors(length(brk)), breaks=lim, 
       xlim=c(0,max_num), right=F, 
       main="Frequency of Daily Steps \n - Missing Values Omitted -", 
       xlab="Steps", cex.axis=0.8)
```

**Mean and Median of dataset "data" with no missing values (in steps)**

```{r}
mean <- mean(data$Steps)
mean

median <- median(data$Steps)
median
```

####                    AVERAGE DAILY ACTIVITY PATTERNS

**1. Time series plot of the 5-minute interval (x-axis) and daily step averages**

```{r}
data <- tapply(act.df$steps, act.df$interval, na.rm = TRUE, mean)
avgSteps <- as.numeric(data)
interval <- as.numeric(levels(factor(act.df$interval)))

plot(interval, avgSteps, type = "l", col = "blue", xlab = "5-Minute Interval", 
        ylab = "Average Number of Steps", main = "Average Daily Activity")
```

**2. 5-minute interval with maximum number of steps**

```{r}
data.df <- data.frame(interval, avgSteps)
data.df[data.df[, 2] == max(avgSteps), ][1]
```  

####                   INPUTTING MISSING VALUES

**1. Calculate and report the total number of missing values in dataset**

```{r}
# Number of missing values in the original dataset:
sum(is.na(act.df))
```

**2. Devise a strategy for filling missing values**

```{r}
#          Mean for 5-minute interval is used as filler

allData <- act.df
for (i in 1:nrow(allData)) {
       if (is.na(allData[i, 1])) 
       allData[i, 1] <- data.df[data.df[, 1] == allData[i, 3], ][, 2]
}
```
**3. Create a dataset identical to the original but with filled in NAs**

```{r}
# New dataset was created that is equal to the original one. 
# Check for missing values in the new dataset:

sum(is.na(allData))
```

**4. Histogram of daily steps and calculation of mean and median**

```{r}
replacedSteps <- aggregate(allData$steps, list(Day=allData$date),
                       na.rm=TRUE, sum)
names(replacedSteps)[2]<-"Steps"
brk<-c(replacedSteps$Day)
lim<-dim(replacedSteps)[1]

hist(replacedSteps$Steps, col=heat.colors(length(brk)), breaks=lim, 
       right=F, main = "Frequency of Daily Steps \n - Missing Values Replaced -", 
       xlab = "Daily Steps",las=1, cex.axis=0.8)
```

**Mean and Median**

```{r}
mean <- mean(replacedSteps$Steps, na.rm = TRUE)
mean

median <- median(replacedSteps$Steps, na.rm = TRUE)
median
summary(replacedSteps$Steps, na.rm=TRUE)
```

####  DIFFERENCE IN ACTIVITY PATTERNS BETWEEN WEEKDAYS AND WEEKENDS

**1. Create a new factor variable with levels: weekday and weekend**

```{r}
allData$weekday <- c("Weekday")
allData[weekdays(as.Date(allData[, 2])) %in% c("Saturday", "Sunday"), 
        ][4] <- c("Weekend")
```

**2. Plot of 5-minute intervals and steps averaged over weekdays/weekends**

```{r}
# Loading necessary packages for time series plot

library(ggplot2)
library(reshape2)
library(plyr)

a<-allData
output<-sapply(split(a, a$weekday), function(a) 
        tapply(a$steps, a$interval, mean))

cleanData <- data.frame(Interval = rep(interval, 2), 
        avgAllSteps = as.vector(unlist(output)), 
        DAY = factor(rep(c("Weekdays", "Weekends") )))

ggplot(cleanData, aes(Interval, avgAllSteps)) + 
        geom_line(aes(colour=DAY)) +
        ggtitle("Time Series Plot of \n Average Steps per Interval") +
        facet_grid(. ~ DAY) + stat_smooth(method = "lm") +
        xlab("5mn Interval") + ylab("Average Steps")
```